### 2013-09-09 background.tex
placeholders

Clinical Background and Significance 
Prostate cancer (PCa) is the most common noncutaneous malignancy among men in the United States. Approximately 1 in every 6 men will develop PCa during their lifetime, with the median age of diagnosis at 67. 1It is also the second leading cause of cancer related death, with 1 in 36 men dying from the disease. The National Cancer Institute estimates that 238,590 men will be diagnosed with PCa in 2013 and 29,720 will die from the disease.1

@@

The Use of Magnetic Resonance Imaging in Prostate Diagnostics
Magnetic resonance (MR) imaging has been available for use in the workup of patients with PCa since the early 1980s but the early work on its diagnostic accuracy is heterogeneous.7 Earlier MR techniques relied mostly on morphology via T1 and T2- weighted imaging (T2WI). The more recent ability to include not only anatomic but also biologic and functional dynamic parameters into MR analysis—via diffusion-weighted imaging (DWI), dynamic contrast-enhanced (DCE) imaging or MR spectroscopic imaging (MRSI)—is promising in the future diagnosis and management of PCA.

@@

 T2-Weighted Imaging and Prostate Anatomy
 T2WI sequences are crucial components of prostate MR imaging.  T2WI is particularly useful in prostate analysis due to its excellent soft tissue contrast resolution, which can be maximized by using thin sections of 3-4mm and a small field of view of approximately 14cm.3,9 T2 sequences are the most helpful for tumor localization as they can clearly show overall prostate morphology, internal structures and prostatic margins.3

@@

 Functional MR Sequences
 Even though T2WI is the mainstay of prostate MR, its overall performance in prostate cancer diagnosis is not optimal. The incorporation of two or more functional sequences in multiparametric MR imaging (mpMRI) has been shown to significantly improve the performance of MRI in cancer diagnosis.12 In fact, the European Society of Urogenital Radiology’s (ESUR) prostate MR guidelines recommend at least 2 functional imaging techniques in addition to T2WI in order to better characterize prostate tumors.8 In a study by Turkbey et al, researchers found that mpMRI had a positive predictive value of 98\% in prostate cancer detection.12 Functional sequences include diffusion- weighted imaging (DWI), dynamic contrast enhanced imaging (DCE-MRI) and MR spectroscopic imaging (MRSI).

 Diffusion-weighted imaging
 DWI is based on the free movement of water particles in tissue and measures the degree of motion restriction.13 Normal prostatic tissue is very glandular with plenty of water molecule movement. On the other hand, tumors have high cellular density and restricted water movement, which leads to decreased diffusion. Due to this restricted diffusion, tumors seem to be of higher intensity on DWI.13

@@

 Dynamic contrast enhanced imaging
 DCE-MRI has been shown to have a high sensitivity in cancer detection, especially when combined with other MR imaging modalities.19-22 In a study by Hara et al, this functional modality identified 93\% of clinically significant cancers.23 DCE-MRI provides a measure of tumor vascularity and takes advantage of tumor-induced angiogenesis in PCa. 24Increased vascularity in PCa results in earlier and higher peak contrast enhancement and a faster washout when compared to normal prostatic tissue. These known characteristics of PCa are helpful when trying to localize lesions.

@@

    MR spectroscopy
    Like DWI, MRSI can be helpful in lesion characterization as it provides information on the presence of certain metabolites in tissue. 8Choline and citrate are especially useful in the setting of PCa. Choline is critical in cell membrane synthesis and is thus usually elevated in cells with cancerous behavior.3 Healthy prostate epithelium synthesizes and secretes large quantities of citrate but levels are decreased in PCa.25,26 Therefore, an increase in the choline-to-citrate ratio on MRSI can be used as an indicator of malignancy.9

>>>>

\subsection{Clinical Background and Significance}
Prostate cancer (PCa) is the most common noncutaneous malignancy among men in the United States. Approximately 1 in every 6 men will develop PCa during their lifetime, with the median age of diagnosis at 67. 1It is also the second leading cause of cancer related death, with 1 in 36 men dying from the disease. The National Cancer Institute estimates that 238,590 men will be diagnosed with PCa in 2013 and 29,720 will die from the disease.1

@@

\subsection{The Use of Magnetic Resonance Imaging in Prostate Diagnostics}
Magnetic resonance (MR) imaging has been available for use in the workup of patients with PCa since the early 1980s but the early work on its diagnostic accuracy is heterogeneous.7 Earlier MR techniques relied mostly on morphology via T1 and T2- weighted imaging (T2WI). The more recent ability to include not only anatomic but also biologic and functional dynamic parameters into MR analysis—via diffusion-weighted imaging (DWI), dynamic contrast-enhanced (DCE) imaging or MR spectroscopic imaging (MRSI)—is promising in the future diagnosis and management of PCA.

@@

\subsubsection{T2-Weighted Imaging and Prostate Anatomy}
 T2WI sequences are crucial components of prostate MR imaging.  T2WI is particularly useful in prostate analysis due to its excellent soft tissue contrast resolution, which can be maximized by using thin sections of 3-4mm and a small field of view of approximately 14cm.3,9 T2 sequences are the most helpful for tumor localization as they can clearly show overall prostate morphology, internal structures and prostatic margins.3

@@

\subsubsection{Functional MR Sequences}
 Even though T2WI is the mainstay of prostate MR, its overall performance in prostate cancer diagnosis is not optimal. The incorporation of two or more functional sequences in multiparametric MR imaging (mpMRI) has been shown to significantly improve the performance of MRI in cancer diagnosis.12 In fact, the European Society of Urogenital Radiology’s (ESUR) prostate MR guidelines recommend at least 2 functional imaging techniques in addition to T2WI in order to better characterize prostate tumors.8 In a study by Turkbey et al, researchers found that mpMRI had a positive predictive value of 98\% in prostate cancer detection.12 Functional sequences include diffusion- weighted imaging (DWI), dynamic contrast enhanced imaging (DCE-MRI) and MR spectroscopic imaging (MRSI).

\subsubsection{Diffusion-weighted imaging}
 DWI is based on the free movement of water particles in tissue and measures the degree of motion restriction.13 Normal prostatic tissue is very glandular with plenty of water molecule movement. On the other hand, tumors have high cellular density and restricted water movement, which leads to decreased diffusion. Due to this restricted diffusion, tumors seem to be of higher intensity on DWI.13

@@

\subsubsection{Dynamic contrast enhanced imaging}
 DCE-MRI has been shown to have a high sensitivity in cancer detection, especially when combined with other MR imaging modalities.19-22 In a study by Hara et al, this functional modality identified 93\% of clinically significant cancers.23 DCE-MRI provides a measure of tumor vascularity and takes advantage of tumor-induced angiogenesis in PCa. 24Increased vascularity in PCa results in earlier and higher peak contrast enhancement and a faster washout when compared to normal prostatic tissue. These known characteristics of PCa are helpful when trying to localize lesions.

@@

\subsubsection{MR spectroscopy}
Like DWI, MRSI can be helpful in lesion characterization as it provides information on the presence of certain metabolites in tissue. 8Choline and citrate are especially useful in the setting of PCa. Choline is critical in cell membrane synthesis and is thus usually elevated in cells with cancerous behavior.3 Healthy prostate epithelium synthesizes and secretes large quantities of citrate but levels are decreased in PCa.25,26 Therefore, an increase in the choline-to-citrate ratio on MRSI can be used as an indicator of malignancy.9
